---
title: "Week 7: Making comparisons between groups and strata"
---

```{r}
week <- 7
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

- Why making comparisons is important
- How to decide on what comparison to make
- Comparing between strata, relative to a baseline
- Comparing the same data, in many ways
- Using normalising to compare different distributions
- Inference using bootstrap and lineups

## Lecture slides

- [html](https://ddde.numbat.space/week7/slides.html)
- [pdf](https://ddde.numbat.space/week7/slides.pdf)
- [qmd](https://ddde.numbat.space/week7/slides.qmd)
- [R](https://ddde.numbat.space/week7/slides.R)

## Tutorial instructions

- [html](https://ddde.numbat.space/week7/tutorial.html)
- [qmd](https://ddde.numbat.space/week7/tutorial.qmd)

```{r}
#| output: asis
show_assignments(7)
```
