---
title: "ETC5521: Diving Deeply into Data Exploration"
---

```{r}
#| label: load_packages
#| include: false
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
options(knitr.kable.NA = '')
source(here::here("course_info.R"))
week <- as.integer(1 + ((Sys.Date() - as_date(start_semester))/7))
```

## Lecturer/Chief Examiner

* **Professor Di Cook**
    - Email: [etc5521.clayton-x@monash.edu](mailto:etc5521.clayton-x@monash.edu) 
    - Consultation: TBD

## Tutors

* Krisanat Anukarnsakulchularp
    - Tutorials: Wed 9:30-11:00am and 7:30-9:00pm CL_Anc-19.LTB_134
    - Consultation: TBD

## Weekly schedule

* Lecture: Tues 10-12 on zoom (link in [Moodle](https://learning.monash.edu/course/view.php?id=18864&section=6))
* Tutorial: 1.5 hours
* Weekly learning quizzes due each Wednesday 9am, from week 2

```{r}
#| label: schedule2
#| message: false
#| warning: false
#| echo: false
#| output: asis
schedule |>
  transmute(
    Show = !is.na(Week), #& (Week <= week | Week <= 1),
    Topic = if_else(!Show, Topic, glue::glue("[{Topic}](./week{Week}/index.html)")),
    Reference = if_else(is.na(Week), Reference, glue::glue("[{Reference}]({Reference_URL})")),
    Assessments = if_else(is.na(Assignment), Assignment, glue::glue("[{Assignment}]({Link})")),
    Week = format(Date, "%d %b")
  ) |>
  select(-Show) |>
  select(Week, everything()) |>
  knitr::kable(format = "markdown")
```

## Assessments

* [Weekly learning quizzes](https://learning.monash.edu/course/view.php?id=9453&section=20): 5%
* Assignment 1: [Instructions](https://learning.monash.edu/mod/assign/view.php?id=2966868) (15%)
* Assignment 2: [Instructions](https://learning.monash.edu/mod/assign/view.php?id=2587917) (20%)
* Assignment 3: [Instructions](https://learning.monash.edu/mod/assign/view.php?id=2587917) (20%)
* Assignment 4 part 1: [Instructions](https://learning.monash.edu/mod/assign/view.php?id=2587918) (20%)
* Assignment 4 part 2: [Instructions](https://learning.monash.edu/mod/assign/view.php?id=2587919) (20%)

## Software

We will be using the latest versions of [R](https://cran.rstudio.com) and [RStudio](https://posit.co/download/rstudio-desktop/). 

Here is the code to install (most of) the R packages we will be using in this unit.

```r
install.packages(c("tidyr", "dplyr", "readr", "readxl", "readabs", "forcats", "tsibble", "cubble", "lubridate", "ggplot2", "GGally", "ggthemes", "sugrrants", "ggbeeswarm", "plotly", "gganimate", "tourr", "sugarbag", "tsibbletalk", "visdat", "inspectdf", "naniar", "validate", "vcd", "mvtnorm", "nullabor", "visage", "forecast", "cassowaryr", "brolgar", "palmerpenguins", "housingData",  "broom", "kableExtra", "lvplot", "colorspace", "patchwork"), dependencies=TRUE)
```

From GitHub, install 

```
remotes::install_github("casperhart/detourr")
```

If you are relatively new to R, working through the materials at [https://learnr.numbat.space](https://learnr.numbat.space) is an excellent way to up-skill. You are epsecially encouraged to work through Chapter 3, on Troubleshooting and asking for help, because at some point you will need help with your coding, and how you go about this matters and impacts the ability of others to help you. 

