---
title: "ETC5521 Worksheet Week 3"
subtitle: "Data validation with pointblank"
author: "Prof. Di Cook"
quarto-required: ">=1.8.0"
format:
    unilur-html:
        output-file: worksheet.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: worksheetsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
        show-solution: true
code-fold: true
---

```{r include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
library(tidyverse)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)

options(
  digits = 2,
  width = 80
)
theme_set(theme_bw(base_size = 14))
```

## 🎯 Objectives

Practice setting up rule-based, automated data validation checks, using the [`pointblank`](https://rstudio.github.io/pointblank/) package, applied to the World Development Indicators (WDI) data from the week 3 lecture. This should take about **30 minutes**.

```{r}
#| eval: false
#| code-fold: false
install.packages(c("tidyverse", "pointblank", "janitor"))
```

## 📦 Data

This is the same tidying used in lecture -- you don't need to change anything here, just run it to get `wdi` (long, one row per country/year/series) and `wdi2016` (a single cross-section).

```{r}
#| code-fold: false
library(janitor)

raw_dat <- read_csv(here::here("data/world-development-indicators.csv"),
                    na = "..", n_max = 11935)

wdi <- raw_dat |>
  select(`Country Code`, `Series Code`, `1969 [YR1969]`:`2018 [YR2018]`) |>
  rename_all(make_clean_names) |>
  pivot_longer(x1969_yr1969:x2018_yr2018,
               names_to = "year",
               values_to = "value") |>
  mutate(year = as.numeric(str_sub(year, 2, 5))) |>
  pivot_wider(names_from = series_code, values_from = value)

wdi2016 <- wdi |> filter(year == 2016)
```

Three series we'll use today:

- `SP.DYN.LE00.IN`: life expectancy at birth, total (years)
- `SP.POP.TOTL`: population, total
- `EN.ATM.CO2E.PC`: CO2 emissions (metric tons per capita) -- note this series stops being reported after 2014, so it will be entirely missing in `wdi2016`.

## 🧩 Tasks

#### 1. Skim the [pointblank introduction](https://rstudio.github.io/pointblank/articles/VALID-I.html). In your own words, what is an `agent`, and why is `interrogate()` a separate step from adding the validation checks?

::: unilur-solution

An `agent` is an object that holds a *plan* of validation checks (a set of rules) for a particular table, built up with `create_agent()` and then piped through validation functions like `col_vals_not_null()`. Nothing is actually checked against the data until you call `interrogate()`, which runs every rule and records the results. Separating the two means you can build up and re-use a validation plan (e.g. save it, apply it to next month's data refresh) without re-running it every time you add a step.

:::

#### 2. Create a pointblank `agent` for `wdi2016`, and add two *structural* checks: `country_code` is never missing (`col_vals_not_null`), and `country_code` is unique (one row per country) (`rows_distinct`). `interrogate()` and check whether `all_passed()`.

::: unilur-solution

```{r}
library(pointblank)

agent <- create_agent(
  tbl = wdi2016,
  tbl_name = "wdi2016",
  label = "WDI 2016 cross-section checks"
) |>
  col_vals_not_null(columns = vars(country_code)) |>
  rows_distinct(columns = vars(country_code)) |>
  interrogate()

all_passed(agent)
```

Both checks pass: every row has a `country_code`, and there's exactly one row per country in this cross-section.

:::

#### 3. Now add two *plausibility* checks: life expectancy (`SP.DYN.LE00.IN`) should be between 0 and 100 years (`col_vals_between`), and population (`SP.POP.TOTL`) should not be negative (`col_vals_gte`). `interrogate()` again. Does the console output say every step is "OK"? Does `all_passed()` agree? If not, why not? (use `get_agent_report` here)

::: unilur-solution

```{r}
agent <- create_agent(
  tbl = wdi2016,
  tbl_name = "wdi2016",
  label = "WDI 2016 cross-section checks"
) |>
  col_vals_not_null(columns = vars(country_code)) |>
  rows_distinct(columns = vars(country_code)) |>
  col_vals_between(columns = vars(SP.DYN.LE00.IN), left = 0, right = 100) |>
  col_vals_gte(columns = vars(SP.POP.TOTL), value = 0) |>
  interrogate()

agent
all_passed(agent)
```

The printed summary shows a ✔ "OK" for every step, but `all_passed()` returns `FALSE`. That's because the console summary only flags steps that breach a *warn/stop* threshold (none are set here), while `all_passed()` asks whether *every single row* passed. Looking at the detailed report shows why:

```{r}
get_agent_report(agent, display_table = FALSE) |>
  select(i, columns, units, n_pass, f_pass)
```

The life expectancy check only passes for 91.7% of rows (18 of 217 countries have a missing value in 2016), and the population check passes for 99.5% (1 country missing). By default, pointblank counts a missing value as a *failing* unit for `col_vals_between()`/`col_vals_gte()`, since it can't confirm the value satisfies the rule.

You can find which countries these are:

```{r}
wdi2016 |> filter(is.na(SP.DYN.LE00.IN)) |> select(country_code)
```

:::

#### 4. We already know from the lecture that this data has missingness, and haven't decided yet whether to treat it as a problem here. Re-run the same two plausibility checks with `na_pass = TRUE`, and explain what that argument changes -- and when you would, or wouldn't, want to use it.

::: unilur-solution

```{r}
agent <- create_agent(
  tbl = wdi2016,
  tbl_name = "wdi2016",
  label = "WDI 2016 cross-section checks"
) |>
  col_vals_not_null(columns = vars(country_code)) |>
  rows_distinct(columns = vars(country_code)) |>
  col_vals_between(columns = vars(SP.DYN.LE00.IN), left = 0, right = 100, na_pass = TRUE) |>
  col_vals_gte(columns = vars(SP.POP.TOTL), value = 0, na_pass = TRUE) |>
  interrogate()

all_passed(agent)
```

`na_pass = TRUE` tells pointblank to *excuse* `NA` values from the check -- a missing value is neither counted as a pass nor a fail, it's simply skipped, so it can no longer drag the pass rate down. Now `all_passed()` is `TRUE`.

Use `na_pass = TRUE` when missingness is already known about and accepted (e.g. you documented it during IDA, and it doesn't itself indicate a data quality problem for this particular check). Leave it `FALSE` (the default) when you want validation to *catch* missingness -- for example, if a future data refresh should always be complete, and any new `NA` appearing would be a red flag worth investigating.

:::

#### 5. *(Bonus, if time permits)* Add a rule for `EN.ATM.CO2E.PC` (CO2 emissions per capita should not be negative) to the agent, using `na_pass = TRUE`, and interrogate. Does it pass? Is that reassuring?

::: unilur-solution

```{r}
agent <- create_agent(tbl = wdi2016, tbl_name = "wdi2016") |>
  col_vals_gte(columns = vars(EN.ATM.CO2E.PC), value = 0, na_pass = TRUE) |>
  interrogate()

get_agent_report(agent, display_table = FALSE) |>
  select(i, columns, units, n_pass, f_pass)
```

It passes 100% -- but not reassuringly so. `EN.ATM.CO2E.PC` stops being reported after 2014, so *every* value in `wdi2016` is `NA`, and with `na_pass = TRUE` all of them are excused rather than checked. A validation step that always "passes" because there's nothing left to check isn't actually telling you anything useful; it's worth checking the *coverage* of a rule (how many non-missing values it actually evaluated), not just whether it passed.

:::

