---
title: "On the relationship between weather and atmospheric pollutants"
author: "YOUR TEAM NAME AND MEMBERS HERE"
date: "today"
format: 
  html:
    css: "assignment.css"
---

```{r}
#| echo: false
# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| echo: false
# Load libraries
library(tidyverse)
library(readxl)
library(visdat)
library(naniar)
library(imputeTS)
library(rvest)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)

# Plot options and themes
options(
  digits = 2,
  width = 60,
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)

theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)
```

## Data description

<!--
Download the daily weather from the gridded data at SILO for the locations of the two sensors used in the first report. This can be down using the weatherOz R package, or directly from the website. You need:

* one year’s worth of data.
* two locations (the ones used by the group you are picking up from in part 1).

plus any additional pollutant data needed to answer the questions. 
-->

## Relationship 1

*Higher temperatures can increase the formation of ground-level ozone* 

## Relationship 2

*Rain and snow can "wash out" particulate matter and water-soluble gases from the air*

## Relationship 3

*High-pressure systems often lead to stagnant air, trapping pollutants.*

## Relationship 4

*Sunlight drives photochemical reactions that form secondary pollutants like ozone.*


## References
