---
title: "Week 6: Bivariate dependencies and relationships, transformations to linearise"
---

```{r}
week <- 6
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

- The humble but powerful scatterplot
- Additions and variations 
- Transformations to linearity
- (Robust) numerical measures of association
- Simpson's paradox
- Making null samples to test for association
- Imputing missing values
- Associations as networks

## Lecture slides

- [html](https://ddde.numbat.space/week6/slides.html)
- [pdf](https://ddde.numbat.space/week6/slides.pdf)
- [qmd](https://ddde.numbat.space/week6/slides.qmd)
- [R](https://ddde.numbat.space/week6/slides.R)

## Worksheet

- [qmd](https://ddde.numbat.space/week6/worksheet.qmd)
- [html](https://ddde.numbat.space/week6/worksheet.html)

## Tutorial instructions

- [html](https://ddde.numbat.space/week6/tutorial.html)
- [qmd](https://ddde.numbat.space/week6/tutorial.qmd)

```{r}
#| output: asis
show_assignments(6)
```
