---
title: "ETC5521 Worksheet 1"
subtitle: "Introduction to exploratory data analysis"
author: "Prof. Di Cook"
quarto-required: ">=1.3.0"
format:
    unilur-html:
        output-file: worksheet.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: worksheetsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
show-solution: true
---

## 🎯 Objectives

The goal of this worksheet is to tackle a data analysis together, by

- mapping out an analysis, with class input and help of AI
- work on cleaning a data set
- making some plots
- discussing what is surprising, and what is expected

## 📋 About the data 

The data to use is available from [Tidy Tuesday 28 May 2024 page](https://github.com/rfordatascience/tidytuesday/blob/master/data/2024/2024-05-28/readme.md). Download the data from here, ideally using the `tidytuesdayR` package. You should only download the data from the Tidy Tuesday once, and save a copy locally on your computer. 

In addition the `gardenR` package, available from `remotes::install_github("llendway/gardenR")` has extra details about the garden. 

::: unilur-solution

This is the code to download the data, and load relevant libraries. 

```{r}
#| eval: false
# remotes::install_github("llendway/gardenR")
# install.packages("tidytuesdayR")
library(tidytuesdayR)
tuesdata <- tidytuesdayR::tt_load('2024-05-28')
save(tuesdata, file="tuesdata.rda")
```

```{r}
library(gardenR)
library(tidyverse)
library(ggbeeswarm)

load("tuesdata.rda")
harvest_2020 <- tuesdata$harvest_2020
harvest_2021 <- tuesdata$harvest_2021
planting_2020 <- tuesdata$planting_2020
planting_2021 <- tuesdata$planting_2021
spending_2020 <- tuesdata$spending_2020
spending_2021 <- tuesdata$spending_2021

data("garden_coords")
```

:::

## 🧩 Tasks 

#### 1. What are the variable types, how and when was the data collected?

::: unilur-solution
:::

#### 2. What would some possible questions be to ask about this data?

::: unilur-solution
:::

#### 3. What do you expect to find?

::: unilur-solution
:::

#### 4. Pick one of the questions, and let's try to answer it. 

- What summaries should we make?
- Are we likely going to need to pre-process the data in any way?

::: unilur-solution
:::
