---
title: "Week 9: Exploring data having a space and time context Part I"
---

```{r}
week <- 9
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

- What is temporal data?
- What is exploratory temporal data analysis?
- Using temporal objects in R: `tsibble`
- Data wrangling: aggregation, creating temporal components, missing values
- Plotting conventions: connect the dots; aspect ratio, landscape or portrait
- Calendar plots: arranging daily records into a calendar format
- Visual inference for temporal data
- tignostics: cognostics for temporal data
- Interactive graphics for temporal data
- Exploring longitudinal data, with the `brolgar` package

## Lecture slides

- [html](https://ddde.numbat.space/week9/slides.html)
- [pdf](https://ddde.numbat.space/week9/slides.pdf)
- [qmd](https://ddde.numbat.space/week9/slides.qmd)
- [R](https://ddde.numbat.space/week9/slides.R)

## Tutorial instructions

- [html](https://ddde.numbat.space/week9/tutorial.html)
- [qmd](https://ddde.numbat.space/week9/tutorial.qmd)

```{r}
#| output: asis
show_assignments(week)
```
