---
title: "Week 10: Exploring data having a space and time context Part II"
---

```{r}
week <- 10
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

- Breaking up data by time, and by space
- Changing focus: 
    - Maps of space over time
    - Exploring time over space with glyph maps
- Inference for spatial trends
- A flash back to the 1970s: Tukey's median polish 
- Working with spatial polygon data
    - Making a choropleth map
    - Bending the choropleth into a cartogram
    - Tiling spatial regions
    
## Lecture slides

- [html](https://ddde.numbat.space/week10/slides.html)
- [pdf](https://ddde.numbat.space/week10/slides.pdf)
- [qmd](https://ddde.numbat.space/week10/slides.qmd)
- [R](https://ddde.numbat.space/week10/slides.R)

## Tutorial instructions

- [html](https://ddde.numbat.space/week10/tutorial.html)
- [qmd](https://ddde.numbat.space/week10/tutorial.qmd)

```{r}
#| output: asis
show_assignments(10)
```
