---
title: "ETC5521: Exploratory Data Analysis"
subtitle: "Initial data analysis and model diagnostics"
author: "Di Cook"
email: "ETC5521.Clayton-x@monash.edu"
date: "Week 3 - Session 2"
length: "50 minutes"
titlebgimg: "images/bg-12.png"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/font-awesome-all.css"
      - "assets/tachyons-addon.css"
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/table.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/slide-types.css"
    self_contained: false 
    seal: false 
    chakra: 'lib/remark-latest.min.js'
    lib_dir: lib
    mathjax: "lib/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: magula
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
current_file <- knitr::current_input()
basename <- gsub(".Rmd$", "", current_file)
```
```{r, include = FALSE}
library(tidyverse)
library(colorspace)
library(patchwork)
options(width = 200)
knitr::opts_chunk$set(
  fig.path = sprintf("images/%s/", basename),
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  fig.retina = 3,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  cache.path = "cache/"
)
theme_set(ggthemes::theme_gdocs(base_size = 18) +
            theme(plot.background = element_blank(), 
                  axis.line.y = element_line(color = "black", linetype = "solid"),
                  plot.title.position = "plot"
                  ) )
```


```{r titleslide, child="assets/titleslide.Rmd"}
```

---
# .circle.bg-black.white[2] Hypothesis Testing and Predictive Modeling.f4[Part 3/3]

.w-60[
- Hypothesis testing: usually make assumptions about the distribution of the data, and are formed relative to a parameter.
- Predictive modeling: form of the relationship, distribution of the errors. 
]

---

# Hypothesis testing in R .f4[REVIEW] .f5[Part 1/3]

.w-90[
- State the hypothesis (pair), e.g. $H_o: \mu_1 = \mu_2$ vs $H_a: \mu_1 < \mu_2$. 
- Test statistic depends on _assumption_ about the distribution, e.g. 
    - $t$-test will assume that distributions are _normal_, or small departures from if we have a large sample. 
    - two-sample might assume both groups have the _same variance_ 
- Steps to complete: 
    - Compute the test statistic
    - Measure it against a standard distribution
    - If it is extreme, $p$-value is small, decision is to reject $H_o$
    - $p$-value is the probability of observing a value as large as this, or large, assuming $H_o$ is true.
]

---
# .blue[Example] .circle.bg-blue.white[1] Checking variance and distribution assumption .f4[Part 1/2]

.pull-left[
.f4[
```{r sleep, echo = TRUE, fig.width=4, out.width="80%"}
data(sleep)
ggplot(sleep, aes(x=group, y=extra)) + 
  geom_boxplot() +
  geom_point(colour="orange")
```
]
]
.pull-right[
```{r echo = TRUE, fig.height=4, fig.width=6, out.width="90%"}
ggplot(sleep, aes(x=extra)) + 
  geom_density(fill="orange", colour="orange", alpha=0.6) + 
  geom_rug(outside = TRUE, colour="orange") +
  coord_cartesian(clip = "off") +
  facet_wrap(~group)
```
]
.footnote[Cushny, A. R. and Peebles, A. R. (1905) The action of optical isomers: II hyoscines. The Journal of Physiology 32, 501–510.]

---
# .blue[Example] .circle.bg-blue.white[1] Hypothesis test .f4[Part 2/2]

.pull-left[
```{r echo = TRUE}
tt <- with(sleep,
     t.test(extra[group == 1],
            extra[group == 2], 
            paired = TRUE))
```

```{r echo = TRUE}
tt$estimate
```

```{r echo = TRUE}
tt$null.value
```
]

.pull-right[
```{r echo = TRUE, results='asis'}
tt$statistic
```

```{r echo = TRUE, results='asis'}
tt$p.value
```

```{r echo = TRUE, results='asis'}
tt$conf.int
```
]
.footnote[Cushny, A. R. and Peebles, A. R. (1905) The action of optical isomers: II hyoscines. The Journal of Physiology 32, 501–510.]

---
# .blue[Example] .circle.bg-blue.white[2] Checking distribution assumption 

.pull-left[
```{r echo=TRUE}
InsectSprays %>% 
  ggplot(aes(x=fct_reorder(spray, count), 
             y=count)) + 
  geom_jitter(width=0.1, height=0, colour="orange", size=3, alpha=0.8) +
  xlab("") 
```

.f4[Can you see any violations of normality? Or equal variance?]

]

.pull-right[

```{r echo=TRUE}
fm1 <- aov(count ~ spray, data = InsectSprays)
summary(fm1)
```

.f4[Write down the hypothesis being tested. What would the decision be?]
]

---


# Linear models in R .f4[REVIEW] .f5[Part 1/3]

.f4[
```{r, echo = TRUE}
library(tidyverse)
library(broom)
glimpse(cars)
```
]
--

.f4[
```{r plot-cars, echo = TRUE, fig.height = 3.5}
ggplot(cars, aes(speed, dist)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```
]


---

# Linear models in R .f4[REVIEW] .f5[Part 2/3]

* We can fit linear models in R with the `lm` function:
```r
lm(dist ~ speed, data = cars)
```
is the same as
```r
lm(dist ~ 1 + speed, data = cars)
```

--

* The above model is mathematically written as 
$$y_i = \beta_0 + \beta_1 x_i + e_i$$
where <ul>
<li>\\(y_i\\) and \\(x_i\\) are the  stopping distance (in ft) and speed (in mph), respectively, of the \\(i\\)-th car;</li> <li>\\(\beta_0\\) and \\(\beta_1\\) are intercept and slope, respectively; and</li>
<li>\\(e_i\\) is the random error; usually assuming \\(e_i \sim NID(0, \sigma^2)\\). </li>
</ul>


---

# Linear models in R .f4[REVIEW] .f5[Part 3/3]

.flex[
.w-20[
```{r plot-cars, fig.align="left", fig.height = 2, fig.width=3}
```

]
.w-80[
```{r, echo = TRUE}
fit <- lm(dist ~ 1 + speed, data = cars)
tidy(fit)
glance(fit)
```
]

]




--

.w-80[
* *Assuming* this model is appropriate, .monash-blue[the stopping distance increases by about `r scales::comma(coef(fit)[2], 1)` ft for increase in speed by 1 mph.]
]


---

count: false

# .circle.bg-black.white[2] Model form .f4[Part 1/2]

.w-60[
* Say, we are interested in characterising the price of the diamond in terms of its carat.
```{r plot-diamonds, fig.height = 5, fig.width = 6}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/5) + 
  ggtitle("Diamonds") #+ 
  #geom_smooth(method = "lm", se = FALSE) + 
  #ylim(0, max(diamonds$price))
```
* Looking at this plot, would you fit a linear model with formula

.center[
`price ~ 1 + carat`?
]

]

---

# .circle.bg-black.white[2] Model form .f4[Part 1/2]

.w-60[
* Say, we are interested in characterising the price of the diamond in terms of its carat.
```{r plot-diamonds-lm, fig.height = 5, fig.width = 6}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/5) + 
  ggtitle("Diamonds") + 
  geom_smooth(method = "lm", se = FALSE, size = 2) + 
  ylim(0, max(diamonds$price))
```
* Looking at this plot, would you fit a linear model with formula

.center[
`price ~ 1 + carat`?
]

]

---

# .circle.bg-black.white[2] Model form .f4[Part 2/2]

.flex[
.w-50[
```{r plot-diamonds-lm2, fig.height = 5, fig.width = 6}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/5) + 
  ggtitle("Diamonds") + 
  geom_smooth(method = lm) +
  geom_smooth(method = lm, 
              formula = y ~ poly(x, 2),
              colour = "orange") +
  ylim(0, max(diamonds$price))
```
]
.w-50[
* What about
<center>
<code style="color: orange">price ~ poly(carat, 2)</code>?
</center>
which is the same as fitting:

$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + e_i.$$

{{content}}
]
]


--

* Should the assumption for error distribution be modified if so?
{{content}}
--


* Should we make some transformation before modelling?
{{content}}
--



* Are there other candidate models?


---

# .circle.bg-black.white[2] Model form .f4[Part 2/2]

* Notice that there was _**no formal statistical inference**_ when trying to determine an appropriate model form.
--


* The goal of the main analysis is to characterise the price of a diamond by its carat. This may involve:
   * formal inference for model selection;
   * justification of the selected "final" model; and
   * fitting the final model.
--


* There may be in fact many, many models considered but discarded at the IDA stage. 
--


* These discarded models are hardly ever reported. Consequently, majority of reported statistics give a distorted view and it's important to remind yourself what might _**not**_ be reported.


---

# Model selection


.blockquote.w-70[
All models are _approximate_ and _tentative_; approximate in the sense that no model is exactly true and tentative in that they may be modified in the light of further data

.pull-right[&mdash;Chatfield (1985)] 

]

<br><br>

--

.blockquote[
All models are wrong but some are useful

.pull-right[&mdash;George Box]

]

---
class: transition

# Model diagnostics

---
# Residuals .f4[1/2]

.flex[
.w-70[
```{r plot-diamonds-resid, fig.height = 8, fig.width = 10, out.width="80%"}
library(broom)
d_fit1 <- lm(price ~ carat, data=diamonds)
d_fit2 <- lm(price ~ poly(carat, 2), data=diamonds)
d_fit3 <- lm(price ~ poly(carat, 3), data=diamonds)
d_fit4 <- lm(price ~ poly(carat, 4), data=diamonds)

d_res1 <- augment(d_fit1, diamonds)
d_res2 <- augment(d_fit2, diamonds)
d_res3 <- augment(d_fit3, diamonds)
d_res4 <- augment(d_fit4, diamonds)
 
r1 <- ggplot(d_res1, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res1[sample(1:nrow(d_res1), 5000),], method = "loess", colour="orange", se=F) +
  scale_y_continuous(".resid", breaks=seq(-20000, 10000, 10000), labels=seq(-20, 10, 10)) +
  ggtitle("Linear")  
r2 <- ggplot(d_res2, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res2[sample(1:nrow(d_res2), 5000),], method = "loess", colour="orange", se=F) +
  scale_y_continuous(".resid", breaks=seq(-20000, 10000, 10000), labels=seq(-20, 10, 10)) +
  ggtitle("Quadratic") 
r3 <- ggplot(d_res3, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res3[sample(1:nrow(d_res3), 5000),], method = "loess", colour="orange", se=F) +
  scale_y_continuous(".resid", breaks=seq(-10000, 30000, 10000), labels=seq(-10, 30, 10)) +
  ggtitle("Cubic") 
r4 <- ggplot(d_res4, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res4[sample(1:nrow(d_res4), 5000),], method = "loess", colour="orange", se=F) +
  scale_y_continuous(".resid", breaks=seq(-10000, 10000, 10000), labels=seq(-10, 10, 10)) +
  ggtitle("Quartic") 

r1 + r2 + r3 + r4 + plot_layout(ncol=2)
```
]
.w-30[
<br>

Residual = Observed - Fitted

<br><br>
Residual plot: Plot the residual against explanatory variable (or Fitted value)

.monash-orange2[Best] residual plot has not .monash-orange2[obvious pattern].
]
]

---
# Alternative approach: linearise relationship

.flex[
.w-50[
```{r plot-diamonds-lm3, fig.height = 5, fig.width = 6}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/5) + 
  ggtitle("Diamonds") + 
  geom_smooth(method = lm) +
  scale_x_sqrt() +
  scale_y_sqrt() + ggtitle("Transform both x, y by sq root")
```
]
.w-50[
```{r plot-diamonds-lm4, fig.height = 5, fig.width = 6}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1/5) + 
  ggtitle("Diamonds") + 
  geom_smooth(method = lm) +
  scale_x_log10() +
  scale_y_log10() + ggtitle("Transform both x, y by log10")
```
]]

The .monash-orange2[log transformation of both variables] linearises the relationship, so that a simple linear model can be used, and also corrects the heteroskedasticity. 

---
# Residuals .f4[2/2]

.flex[
.w-70[
```{r plot-diamonds-resid2, fig.height = 8, fig.width = 10, out.width="80%"}
library(broom)
diamonds <- diamonds %>%
  mutate(sqprice = sqrt(price),
         sqcarat = sqrt(carat),
         lprice = log10(price),
         lcarat = log10(carat))
d_fit5 <- lm(sqprice ~ sqcarat, data=diamonds)
d_fit6 <- lm(lprice ~ lcarat, data=diamonds)

d_res5 <- augment(d_fit5, diamonds)
d_res6 <- augment(d_fit6, diamonds)

r5 <- ggplot(d_res5, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res5[sample(1:nrow(d_res5), 5000),], method = "loess", colour="orange", se=F) +
  ggtitle("Square root")  
r6 <- ggplot(d_res6, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res6[sample(1:nrow(d_res6), 5000),], method = "loess", colour="orange", se=F) +
  ggtitle("Log10") 
r7 <- ggplot(d_res5, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res5[sample(1:nrow(d_res5), 5000),], method = "loess", colour="orange", se=F) +
  xlim(c(0, 3)) +
  ggtitle("Square root")  
r8 <- ggplot(d_res6, aes(carat, .resid)) + 
  geom_hline(yintercept=0, colour="grey70") +
  geom_point(alpha = 1/5) +
  geom_smooth(data=d_res6[sample(1:nrow(d_res6), 5000),], method = "loess", colour="orange", se=F) +
  xlim(c(0, 3)) +
  ggtitle("Log10") 

r5 + r6 + r7 + r8 + plot_layout(ncol=2)
```
]
.w-30[
<br>

Which has the best residual plot? 

]
]


---

class: middle center

.blockquote[
"Teaching of Statistics should provide a more balanced blend of IDA and inference"

.pull-right[Chatfield (1985)]
]


--

<br><br>

Yet there is still very little emphasis of it in teaching and also at times in practice.

--

<br>

So don't forget to do IDA!

---

# Take away messages

.flex[
.w-90.f2[



<ul class="fa-ul">
{{content}}
</ul>


]
]

--

  <li><span class="fa-li"><i class="fas fa-paper-plane"></i></span><b><i>Initial data analysis</i></b> (IDA) is a model-focused exploration to support a confirmatory analysis with:
  <ul>
  <li><b><i> data description and collection</i></b> </li>
  <li><b><i> data quality checking, and</i></b></li>
  <li><b><i> checking assumptions</i></b> </li>
  <li><b><i> model fit</i></b> without any formal statistical inference.</li>
  </ul>
</li>

{{content}}

--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>IDA may never see the limelight BUT it forms the foundation that the main analysis is built upon. Do it well!</li>


---

```{r endslide, child="assets/endslide.Rmd"}
```

<br><br>Lecture materials originally developed by Dr Emi Tanaka
