---
title: "ETC5521 Tutorial 3"
subtitle: "Initial data analysis"
author: "Prof. Di Cook"
quarto-required: ">=1.3.0"
format:
    unilur-html:
        output-file: tutorial.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: tutorialsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
        show-solution: true
---


```{r include=FALSE}
#| echo: false
library(tidyverse)
library(ggbeeswarm)
library(nullabor)
library(broom)
library(conflicted)

# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
options(
  digits = 2,
  width = 80,
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)

```

## 🎯 Objectives

Practice conducting initial data analyses, and make a start on learning how to assess significance of patterns. 

## 🔧 Preparation 

The reading for this week is [The initial examination of data](https://www.jstor.org/stable/28c1953a-8130-3c31-bd91-3ca6ce1d96d8?seq=7). It is authored by Chris Chatfield, and is a classic paper explaining the role of initial data analysis. 

- Complete the weekly quiz, before the deadline!
- Open your RStudio Project for this unit, (the one you created in week 1, `ETC5521`). Create a `.qmd` document for this weeks activities. 


## 📥 Exercises

Today, you have sprint competition to discover as many problems as possible in this data, [cafe.rda](https://github.com/numbats/ddde/blob/main/data/cafe.rda).

*A small cafe in the city of Melbourne is interested in determining whether the daily earnings depend on the weather. They compiled data for a period over 2000-2001 to study this question.* The data has the following variables:

```{r}
#| echo: false
library(kableExtra)
desc <- tibble(var=c("dt", "wday", "revenue", "expend", "precip", "mint", "maxt", "source"),
               description=c("Date", 
              "Day of the week",
              "Daily revenue in hundreds, 11=1100", 
              "Daily expenses in hundreds",
              "Precipitation in mm", 
              "Minimum temperature, Celsius",
              "Maximum temperature, Celsius",
              "Source of the weather data"))
kbl(desc, table.attr = 'data-quarto-disable-processing="true"') |> 
  kable_styling(full_width = FALSE) |>
  column_spec(1, width="2cm", border_right = T) |>
  column_spec(2, width="10cm")
```

#### Guidelines

- Use whatever R package or software or tool you'd like, and report how you found the error.
- Feel free to buddy up, and work with another student in your tutorial session.
- No cheating! This time do the assignment without AI help, or searching for answers on the web.
- Please don't ruin the later tutorials by sharing your work!

The most complete list with best process for findings, as decided by your tutor, wins the prize! Your tutor's decision is final!

::: unilur-solution
Your tutor has the solution!
:::

## 👌 Finishing up

Make sure you say thanks and good-bye to your tutor. This is a time to also report what you enjoyed and what you found difficult.
