---
title: "Week 8: Going beyond two variables, exploring high dimensions"
---

```{r}
week <- 8
source(here::here("course_info.R"))
```

## Main reference

```{r}
#| output: asis
schedule |>
  filter(Week == week) |>
  mutate(ref = glue::glue("[{Reference}]({Reference_URL})")) |>
  pull(ref)
```

## What you will learn this week

- What is high-dimensional data? (If all variables are quantitative)
- Exploring relationships between more than two variables
    - Tours - scatterplots of combinations of variables
    - Matrix of plots
    - Parallel coordinates
- What can be hidden
- Automating the search for pairwise relationships using scagnostics
- Linking elements of multiple plots
- Exploring multiple categorical variables

## Lecture slides

- [html](https://ddde.numbat.space/week8/slides.html)
- [pdf](https://ddde.numbat.space/week8/slides.pdf)
- [qmd](https://ddde.numbat.space/week8/slides.qmd)
- [R](https://ddde.numbat.space/week8/slides.R)

## Tutorial instructions

- [html](https://ddde.numbat.space/week8/tutorial.html)
- [qmd](https://ddde.numbat.space/week8/tutorial.qmd)

```{r}
#| output: asis
show_assignments(week)
```
